Model for DES Type Ia Supernova Cosmology Analysis

### 2016-05-20 tex/toy.tex
redo efficiencies in more rigorous fashion

The likelihood is related to the pdf of observables given the model for the population of transients that satisfy sample
selection criteria that is different from the population of all transients: we thus label the likelihood with
an $S$ subscript to emphasize that its population is distinct from the underlying population.
The likelihood for observations $O_o$ is expressed as
\begin{align}
\mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where $p(O|M, M_I)$ is the pdf of the underlying sample, and $\epsilon(O, M, M_I) $
is the probability those observables will be included in the analysis sample.
As the observable uncertainties across different transients
are expected to be independent the likelihood can be split into contributions
from each,
\begin{align*}
\mathcal{L}_S(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients} \epsilon(O_o^\alpha, M, M_I) p(O_o^\alpha|M, M_I)}{\int \prod_{\alpha \in transients}  \epsilon(O^\alpha, M, M_I) p(O^\alpha|M, M_I)dO^\alpha}.
\end{align*}
Note that we work with counts:  correlated calibration uncertainties would prevent this splitting for fluxes.

@@

The efficiency $\epsilon(O, M, M_I) $ depends on the properties of the experiment and the sample selection criteria:
the transient is discovered, satisfies sample selection, is selected for spectroscopic typing, has successful typing, and is SN~Ia:
\begin{align*}
\epsilon(O, M, M_I)  & = P(SNIa, typed, spec, sample, discovered |O, M, L, \tau)\\
& = P(SNIa|  typed, \tau)P(typed | spec, s, M, M_I)P(spec | discovered, c) P(sample, discovered| c).
\end{align*}
Transient discovery and sample selection are purely based on observed photometry $c$.
The decision for spectroscopic follow-up is also based on observed photometry,  though perhaps external redshift estimates
of the possible host galaxy may come into play.  The success of the typing measurement should primarily depend
on the signal $s$, but may also depend on the type and host properties.

A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as follows:
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}_S(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}_S(O_o| M, M_I) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}
Once all the $p$'s and $P$'s are specified we are ready to analyze!

@@

\subsection{$p(M_I|M)$: Probabilities of the Intermediary Parameters.}
\begin{itemize}
\item Type probabilities are relevant in the photometric analysis.
The probability an object is SN~Ia is
\begin{align}
p_{SNIa} &= \frac{r_{SNIa}}{r_{SNIa}+r_{SNII}} \nonumber \\
p_{SNII}&=1-p_{SNIa}.
\label{prob:eqn}
\end{align}

@@

\subsection{$p(O| M, M_I)$: Probability of All Possible Observables}
\begin{itemize}
\item The spectroscopic typing is assumed to be perfect
\begin{equation}
T | \tau \sim \delta (T-\tau).
\end{equation}

@@

\subsection{$\epsilon(O, M, M_I)$: Sample Inclusion Efficiency}
The probability that a transient becomes a part of the analysis sample is given by $\epsilon(O, M, M_I)$ and is dependent
on the properties of the experiment and choices made by the experimenter.
\begin{itemize}
\item The probability the transient is discovered and included in the sample is given by
$P(sample, discovered |c)$.  To first order, the discovery probability within a single image depends on the transient signal-to-noise $\epsilon(STON)$.  For
an image with noise $N$ the efficiency for discovery is $\epsilon(c/N)$, and the probability of discovery in at least one image
\begin{equation}
\epsilon_D(c) = 1-\prod_{\beta \in images} \left(1-\epsilon(c_\beta/N_\beta)\right).
 \end{equation}

As an illustrative example we write out the probability of a sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$:
\begin{align*}
\epsilon_S(c) = &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c)\\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \mathcal{N}(c_\star-c_\alpha, C_c)\prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c),
 \end{align*}
 where $\mathcal{H}$ is the Heaviside step function (other sample selection cases are similar but with different combinatorics).
Then 
$P(sample, discovered |c) = \epsilon_D(c) \epsilon_S(c)$.

We could write
\begin{align*}
P(sample, discovered |c) & =P(sample| discovered, c)P(discovered |c)\\
&= \epsilon_S(c)
\end{align*}
if the sample selection criteria are selected such that $P(discovered | sample, c)=1$.
This would have significant
computational benefits in the evaluation of the denominator of Eqn.~\ref{likelihood:eqn}.
\item The probability of obtaining a spectroscopic observation of a transient is given as $P(spec | discovered, c)$.  Each follow-up telescope
has its own criteria: OzDES is magnitude limited, VLT prefers transients on faint hosts, others may try to get a random sample uniform in
redshift.  Mat Smith and Chris D'Andrea should be able to provide this term.
\item The probability that the spectroscopic observation yields a type is $P(typed | spec, s, M, M_I)$.  This is a function we have to
figure out!
\item This term appears for the typed SN~Ia sample, where it is trivial $P(SNIa|  typed, \tau)=1$.
\end{itemize}

%\section{Implementation Notes}
%Anticipating the use of available statistical packages, the likelihoods have been constructed
%from standard distribution functions.  The truncated distributions from sample selection
%are accounted by the normalization 
%terms $P(S_c, S_T| z=z_o, X)$, $P(S_c| z=z_o, X)$.  Although they do not appear
%in the PGM, they are treated as nodes who contribute 
%$-\ln{P(S_c, S_T| z=z_o, X)}$ to the log-likelihood, where the negative sign reflects that this is a normalization
%factor.
%
%
%The following integral appears in the normalization
%\begin{equation}
%\int dL\, p(L|T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L, z=z_o, X)\right].
%\end{equation}
%We actually work with $\ln{L_X}$ and $\sigma_X$.  Setting
%\begin{equation}
%n=\frac{\ln{L}-\ln{L_X}}{\sigma_X}
%\end{equation}
%the integral becomes
% \begin{equation}
%\sigma_X \int dn\, p(n\sigma_X + \ln{L_X} |T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o\,  p(c_o | T=T_i, L_Xe^{n\sigma_X}, z=z_o, X)\right].
%\end{equation}
%The two probability distributions are normal so that this reduces to
% \begin{equation}
%\frac{1}{2\sqrt{2\pi}} \int dn\, e^{-0.5n^2} \left(1-erf\left(\frac{(c_T - c_\star)}{\sqrt{2}\sigma_c}\right)\right),
%\end{equation}
%where
%\begin{equation}
%c_\star = \frac{e^{\ln{L_X}+n\sigma_X}}{4\pi d_L^2}10^{Z/2.5}.
%\end{equation}

>>>>

The model describes transients and their observables.  We are particularly interested in the pdf of those observables
given the model.  The pdf is dependent on the sample and we consider two such samples: an underlying sample
that represents the distribution of all possible transients and observables accommodated by the model; and
an analysis sample that represents the distribution of transients and observables that end up being analyzed.
The analysis sample differs from the underlying sample in that the transients of the former 
must be discovered, pass photometric selection criteria, and for the SN~Ia sample be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.
The pdf of the underlying sample is denoted as $p(O|M,M_I)$ whereas the pdf of the analysis sample, the likelihood,
is denoted $\mathcal{L}_S(O | M, M_I)$. 
The two are related by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}(O_o | M, M_I) & = \frac{p_S(O_o|M, M_I)}{\int  p_S(O|M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I).
\end{align}

As the sample-inclusion probabilities and the uncertainties in $O_o$ are expected to be uncorrelated between different transients
 the likelihood can be split into contributions
from each,
\begin{align*}
\mathcal{L}(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients} p_S(O_o^\alpha|M, M_I)}{\prod_{\alpha \in transients} \int  p_S(O^\alpha|M, M_I)dO^\alpha}.
\end{align*}
Note that we work with counts:  correlated calibration uncertainties would prevent this splitting for fluxes.

@@

The probability density for an object to enter the analysis sample can be broken out
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I)\\
&= P(T, T=SNIa| s, c, disc=True, anal=True, sobs=True, type=True, M, M_I)\\
& \times  p(s, type=True | c, disc=True, anal=True, sobs=True, M, M_I)\\
& \times  p(c, disc=True, anal=True, sobs=True| M, M_I)\\
&= P(T, T=SNIa| M, M_I)  p(s, type=True | M, M_I) \\
& \times  p(c, disc=True, anal=True, sobs=True| M, M_I).
\end{align}

%Transient discovery and sample selection are purely based on observed photometry $c$.
%The decision for spectroscopic follow-up is also based on observed photometry,  though perhaps external redshift estimates
%of the possible host galaxy may come into play.  The success of the typing measurement should primarily depend
%on the signal $s$, but may also depend on the type and host properties.

A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as follows:
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}(O_o| M, M_I) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}
Once all the $p$'s and $P$'s are specified we are ready to analyze!

@@

\subsection{$p(M_I|M)$: Probabilities of the Intermediary Parameters.}
\begin{itemize}
\item Type probabilities are relevant in the photometric analysis.
The probability an object is SN~Ia is
\begin{align}
P_{SNIa} &= \frac{r_{SNIa}}{r_{SNIa}+r_{SNII}} \nonumber \\
P_{SNII}&=1-P_{SNIa}.
\label{prob:eqn}
\end{align}

@@

\subsection{$p(O| M, M_I)$: Probability of All Possible Observables}
\begin{itemize}
\item The spectroscopic typing is assumed to be perfect
\begin{equation}
T | \tau \sim \delta (T-\tau).
\label{typepdf:eqn}
\end{equation}

@@

\subsection{$p_S(O|M, M_I)$: Sample Selection Probability}
We have listed the probabilities for the underlying distribution.  We now describe the probabilities for making it into the analysis sample.
There are three terms.

\begin{itemize}
\item $P(T, T=SNIa| M, M_I)$: the probability a type measurement $T$ yields an SN~Ia type.  It depends
on rate parameters as passed through in Eqn.~\ref{typepdf:eqn}.
\item $p(s, type=True | M, M_I)$: the probability of a spectral measurement $s$ that yields a successful typing.
\item $p(c, disc=True, anal=True, sobs=True| M, M_I)$
\end{itemize}
%\subsection{$\epsilon(O, M, M_I)$: Sample Inclusion Efficiency}
%The probability that a transient becomes a part of the analysis sample is given by  \begin{equation}
%P(typed, spec, sample, discovered | s, c, M, M_I) = \epsilon_T(s, M, M_I) \epsilon_O(c) \epsilon_S(c) \epsilon_D(c)
%\end{equation}
%and is dependent
%on the properties of the experiment and choices made by the experimenter.
%The probability of typing depends on the spectroscopic data whereas the probability of spectroscopic observation,
%sample selection, and discovery depend on photometry.
%
%\begin{itemize}
%\item The probability the transient is discovered 
%in a single image depends on the transient signal-to-noise $\epsilon(STON)$.  For
%an image with noise $N$ the efficiency for discovery is $\epsilon(c/N)$, and the probability of discovery in at least one image
%\begin{equation}
%\epsilon_D(c) = 1-\prod_{\beta \in images} \left(1-\epsilon(c_\beta/N_\beta)\right).
% \end{equation}
%\item
%As an illustrative example we write out the probability of a sample selection requiring two photometric points above
%a signal-to-noise threshold $STON_T$:
%\begin{align*}
%\epsilon_S(c) = &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c)\\
%&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \mathcal{N}(c_\star-c_\alpha, C_c)\prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\mathcal{N}(c_\star-c_\beta, C_c),
% \end{align*}
% where $\mathcal{H}$ is the Heaviside step function (other sample selection cases are similar but with different combinatorics).
%\item The probability of obtaining a spectroscopic observation of a transient is given as $\epsilon_O(c)$.  Each follow-up telescope
%has its own criteria: OzDES is magnitude limited, VLT prefers transients on faint hosts, others may try to get a random sample uniform in
%redshift.  Mat Smith and Chris D'Andrea should be able to provide this term.
%\item The probability that the spectroscopic observation yields a type is $\epsilon_T( s, M, M_I)$.  This is a function we have to
%figure out!
%\end{itemize}

\subsection{$\int  p_S(O|M, M_I)dO$: Normalizing the Likelihood}
We could write
\begin{align*}
P(sample, discovered |c) & =P(sample| discovered, c)P(discovered |c)\\
&= \epsilon_S(c)
\end{align*}
if the sample selection criteria are selected such that $P(discovered | sample, c)=1$.
This would have significant
computational benefits in the evaluation of the denominator of Eqn.~\ref{likelihood:eqn}.

\section{Implementation Notes}
Anticipating the use of available statistical packages, the likelihoods have been constructed
from standard distribution functions.  The truncated distributions from sample selection
are accounted by the normalization 
terms $P(S_c, S_T| z=z_o, X)$, $P(S_c| z=z_o, X)$.  Although they do not appear
in the PGM, they are treated as nodes who contribute 
$-\ln{P(S_c, S_T| z=z_o, X)}$ to the log-likelihood, where the negative sign reflects that this is a normalization
factor.

The following integral appears in the normalization
\begin{equation}
\int dL\, p(L|T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o  p(c_o | T=T_i, L, z=z_o, X)\right].
\end{equation}
We actually work with $\ln{L_X}$ and $\sigma_X$.  Setting
\begin{equation}
n=\frac{\ln{L}-\ln{L_X}}{\sigma_X}
\end{equation}
the integral becomes
 \begin{equation}
\sigma_X \int dn\, p(n\sigma_X + \ln{L_X} |T_i, z=z_o, X)  \left[\int_{c_T}^{\infty} dc_o\,  p(c_o | T=T_i, L_Xe^{n\sigma_X}, z=z_o, X)\right].
\end{equation}
The two probability distributions are normal so that this reduces to
 \begin{equation}
\frac{1}{2\sqrt{2\pi}} \int dn\, e^{-0.5n^2} \left(1-erf\left(\frac{(c_T - c_\star)}{\sqrt{2}\sigma_c}\right)\right),
\end{equation}
where
\begin{equation}
c_\star = \frac{e^{\ln{L_X}+n\sigma_X}}{4\pi d_L^2}10^{Z/2.5}.
\end{equation}
